Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-12-17 paper.tex
take out regression

The computerized tutor was originally built to test the ACT and ACT-R
theories of skill acquisition \citep{anderson2013architecture}.
These are elaborate theories describing the necessary components of
cognitive skills in, say, mathematics or computer programming, and the
process of acquiring those skills.
The Algebra I tutor guides students through a sequence of Algebra I
problems nested in sections within units, and organized by the specific sets of skills they require.
Students move through the curriculum as the tutor's internal learning
model determines that they have mastered the requisite skills.

As students work through specific problems, they receive two types
of instruction from the tutor.
When they err---that is, depart from one of the pre-specified correct
solution paths for that problem---they receive an error message
alerting them to their mistake and helping them return to the correct
solution path.
Students may also request help, in the form of hints, that are also
tailored to the specific context and solution path they are on.

Since our goal was to better understand the CTA1
treatment effect, we focused our analysis on data from high school students in
the second year of the CTA1 trial, for whom the treatment effect was
most evident.

We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
Table \ref{tab:covariateBalance} describes the covariates we used, including
missingness information, control and treatment means, and standardized
differences \citep[c.f.][]{kalton1968standardization}.
We singly-imputed missing values with the Random Forest routine implemented
by the \texttt{missForest} package in \texttt{R}
\citep{missForest,rcite},  which estimated the ``out of box'' imputation
errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.

Over the course of the effectiveness study, Carnegie Learning gathered
log data for each worked problem.
The resulting dataset lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
Treatment schools with mastery data missing
for 90\% or more students were omitted from the analysis, along with
their entire randomization block.
Of the remaining 2390 students, 87\% (2,091) had log
data.
For the sake of simplicity, the treatment students without log data
were dropped from the study.\footnote{Including these students in a
  principal stratification model is straightforward \citep{aoas}; see
  the online supplement for results from a model including them, which
  were broadly similar to the results presented here. Including
  subjects with missing log data in a mediation or observational study
  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
Problem log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

@@

\subsection{Measuring Hint Usage in the CTA1 Data}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package, including the associated textbook and
recommended classroom practices.
Access to the full CTA1 curriculum was randomized, rather than any
specific aspect of usage, and effectiveness was measured with a
standardized posttest covering a broad range of Algebra I skills.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a posttest.

The CTA1 study was of a much longer duration than
any of the studies of hint effects we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.
What is relevant for our purposes is that we may narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.

Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
% (More mundanely, we do not have access to data
% describing the number of hints available for every problem, which
% would hinder interpretation of any other measure of hint usage.)

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately 30\% of students as
high-hint users, so we chose $c$ as the
70th percentile of $\bar{h}$,
approximately 0.6.
This definition agreed with the model's classification in
approximately 93\% of cases.

@@

Let $m_i$ be a categorical variable denoting the matched set that $i$
belongs to.
Then we may re-state the ignorability condition with reference to this
match \citep[c.f.][]{rebar}:
\begin{ass}{Matched Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |m
\end{equation*}
\end{ass}
% Matched ignorability formalizes a ``latent experiment'' design: by
% assuming matched ignorability, we imagine a randomized trial in which
% members of the CTA1 group were divided into strata defined by $m$, and
% randomized to $H=1$ or $H=0$ within those strata.
Under matched ignorability, the mean of the outcomes between $H=1$ and
$H=0$ students within each match is unbiased for the average treatment
effect for students in that match.
The (weighted) mean of those estimated treatment effects estimates the
overall treatment effect in the sample.

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest (the standard error
and confidence interval using the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Oct 28 20:04:28 2019
\begin{table}[ht]
\centering
\begin{tabular}{llrrrrr}
  \hline
  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
  \hline
\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
  \hline &  &  &  &  &  &  \\ 
  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
   \hline
\end{tabular}
\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
30\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
70th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
individual level, as $\delta(1)_i$ and $\xi(0)_i$.
However, as in the usual causal inference case, individual effects are
not identified without strong assumptions, and the target of
estimation is instead taken to be super-population average effects,
$\EE[\delta(1)]$ and $\EE[\xi(0)]$. Our approach to estimating these
is to estimate the means of each of the component potential outcomes,
$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
Y(0,H(0))=Y(Z=0)$ are non-parametrically identified due to the
randomization of treatment assignment.

On the other hand, randomization alone does not identify the strictly counterfactual
potential outcome $\EE Y(1,H(0))$.
In a typical mediation analysis of an RCT, identification of the
distribution of $H(0)$ in the population would derive from
randomization, along with an observation of $H$ in the control group.
In contrast, in the CTA1 study, $H(0)$ is not observed but is
determined by the experimental design: since the CTA1 software was
unavailable to the control group, $H(0)=0$ for all experimental
participants.
Therefore, $Y(1,H(0))\equiv Y(1,0)$.
For treated subjects $i$ for whom $H_i=0$, then, $Y_i(1,H(0))=Y_i$,
the observed outcome is equal to the counterfactual outcome.
For these subjects, there is no indirect effect.

Treated subjects with $H=1$ present more of a challenge.
To estimate indirect effects for these subjects, we may rely on the
same causal setup as in the observational approach:
matched ignorability and the matching design of section
\ref{sec:observational}.
If $Y_i(1,H(0))$ is independent of $Z_i$ within matched sets, then we
may estimate counterfactual potential outcomes for $H=1$ subjects
using the observed outcomes of their matches with $H=0$.
Specifically, let
\begin{equation*}
\hat{Y}_i(1,H(0))=\begin{cases}
 Y_i & H_i=0\\[3ex]
\frac{\displaystyle\sum_{j:M_j=M_i}
  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
\end{cases}
\end{equation*}
For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
average outcome of their matched pairs with $H=0$.
Then, the average indirect effect in the treatment group can be
estimated as:
\begin{equation*}
 \hat{\Delta}\equiv\widehat{\EE[\delta(0)|Z=1]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
\end{equation*}
(Typically, the overall average indirect effect, $\EE[\delta(0)]$, is
estimated instead of the average indirect effect of the treatment
group, $\EE[\delta(0)|Z=1]$.
We estimated the latter since data from the control group was not used
in our estimation strategy, and estimating the overall average would
simply complicate our standard error estimation.)

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=$0.7,
$\hat{\Delta}=$0.7$TOT$.
Estimated thusly, the indirect effect is negative: -0.1, with
a standard error of 0.04.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
0.14$\pm$0.23.
Including covariates in the regression produces a direct effect of
0.18$\pm$0.19.

>>>>

The computerized tutor was originally built to test the ACT and ACT-R
theories of cognition \citep{anderson2013architecture}.
These are elaborate theories that describe, among other things, the necessary components of
cognitive skills in, say, mathematics or computer programming, and the
process of acquiring those skills.
The Algebra I tutor guides students through a sequence of Algebra I
problems nested in sections within units, and organized by the specific sets of skills they require.
Students move through the curriculum as the tutor's internal learning
model determines that they have mastered the requisite skills.

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through a problem---that is, departs from one of the
solution paths---the
software shows the student an error message.
This feedback is tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

@@

\subsection{Measuring Hint Usage in the CTA1 Data}
\marginpar{make sure this makes sense before ``Data'' section}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package, including the associated textbook and
recommended classroom practices.
Access to the full CTA1 curriculum was randomized, rather than any
specific aspect of usage, and effectiveness was measured with a
standardized posttest covering a broad range of Algebra I skills.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a posttest.

The CTA1 study was of a much longer duration than
any of the studies of hint effects we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.

For the current purpose, we narrow down the type
of hint data we model to align with our research focus.
Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set to one if the student requested any
hints on that problem, and zero otherwise.
% (More mundanely, we do not have access to data
% describing the number of hints available for every problem, which
% would hinder interpretation of any other measure of hint usage.)

@@

\subsubsection{Dichotomizing Hint Usage}\label{sec:dichotomizingHintUsage}
The observational study and mediation analysis below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.

To choose $c$, we first fit a modified Rasch mixture model
\citep{rasch} to hint request data.
Specifically, we modeled the probability of a hint request in each problem
$p$, worked by student $i$, as
\begin{equation*}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation*}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems within each section are equally
likely to elicit a hint.

Then, in order to dichotomize hint request behavior, we modeled
student effects $\eta$ as
\begin{equation}\label{eq:mixture}
\eta\sim
p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
\end{equation}
and constrained $\mu_0<\mu_1$.
This mixture model clusters students into two categories: high hint
users have $\eta$ drawn from a normal distribution with mean $\mu_1$
and standard deviation $\sigma_1$, and low hint users have $\eta$
drawn from a normal distribution with mean $\mu_0$ and standard
deviation $\sigma_0$.

Our main parameter of interest here is $p_0$, the proportion of
students who are low hint users.
We estimated $p_0\approx$0.7,
classifying 30\% of students as
high hint users, so we chose $c$ as the
70th percentile of $\bar{h}$,
approximately 0.6.
This definition agreed with the model's classification (based on
$\eta$ as opposed to $\bar{h}$ in
approximately 93\% of cases.

The Rasch model for hint useage \eqref{eq:rasch}, but not the mixture
model \eqref{eq:mixture} will play an important role in the principal
stratification analysis of Section \ref{sec:principalStratification}.

\citep{pane2014effectiveness} estimated effects separately in middle
schools and high schools in each the first two years of the RCT; the
only statistically significant treatment effect was in the high school
sample in the second year.
Since our goal was to better understand the CTA1
treatment effect, we focused our analysis on data from high school students in
the second year of the CTA1 trial, for whom the treatment effect was
most evident.

We merged data from two sources: computerized log data gathered by
Carnegie Learning, and covariate, treatment and outcome data gathered by RAND.

The log data lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files
were not retrievable, or because of an imperfect ability to link log
data to other student records.
Further, log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

To construct our analysis dataset, we first dropped treatment schools with log data missing
for 90\% or more students.
Prior to randomization, schools were stratified into pairs or triples
based on baseline covariates, and randomized between treatment and
control condition within these randomization blocks.
When we dropped a treatment school from our analysis, we also dropped
the control schools in its randomization block.
Of the remaining 2,390 students, 87\% (2,091) had log
data.
Then, for the sake of simplicity, the treatment students without log data
were dropped from the study.\footnote{Including these students in a
  principal stratification model is straightforward \citep{aoas}; see
  the online supplement for results from a model including them, which
  were broadly similar to the results presented here. Including
  subjects with missing log data in a mediation or observational study
  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}

Table \ref{tab:covariateBalance} describes the covariates we used, including
missingness information, control and treatment means, and standardized
differences \citep[c.f.][]{kalton1968standardization} from the final
analysis sample.
We singly-imputed missing values with the Random Forest routine implemented
by the \texttt{missForest} package in \texttt{R}
\citep{missForest,rcite},  which estimated ``out of box'' imputation
error rates as part of the random forest regression, also shown in Table \ref{tab:covariateBalance}.

@@

\subsection{Estimating the Effects of $H$ on Posttest Scores}

Let $M_i$ be a categorical variable denoting the matched set that $i$
belongs to.
Then we may re-state the ignorability condition with reference to this
match \citep[c.f.][]{rebar}:
\begin{ass}{Matched Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |M
\end{equation*}
\end{ass}
% Matched ignorability formalizes a ``latent experiment'' design: by
% assuming matched ignorability, we imagine a randomized trial in which
% members of the CTA1 group were divided into strata defined by $m$, and
% randomized to $H=1$ or $H=0$ within those strata.
Under matched ignorability, the mean of the outcomes between $H=1$ and
$H=0$ students within each match is unbiased for the average treatment
effect for students in that match.
In other words, if $\tau_m=\overline{Y(H=1)}-\overline{Y(H=0)}$ is the
average effect in match $m$, then
$\hat{\tau}_m=\bar{Y}_{M=m,H=1}-\bar{Y}_{M=m,H=0}$ is an unbiased
estimate of $\tau_m$, where $\bar{Y}_{M=m,H=h}$ is the average
observed outcome for subjects in match $m$ with hint level
$h\in\{0,1\}$.

Weighted means of those estimated treatment effects, of the form
\begin{equation}\label{eq:tauWeightedMean}
\hat{\tau}_w=\sum_m w_m\hat{\tau}_m/\sum_m w_m
\end{equation}
estimate aggregate
treatment effects in the sample.
To estimate the average effect of $H$ for all subjects, set $w_m=
n_{1m}+n_{0m}$, where $n_{1m}$ and $n_{0m}$ are the number of
$H=1$ and $H=0$ subjects in match $m$, respectively.
Alternatively, setting $w_m=n_{1m}$ estimates the average effect of
$H$ on those subjects for whom $H=1$---the ``treatment on the
treated,'' or TOT, effect.
Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$, the harmonic mean  are
``precision weights'' \citep[c.f.][]{schochet2015statistical}; the
estimate using these weights is equal to the coefficient on $H$ from
an ordinary least squares (OLS) regression of $Y$ on $H$ and a set of dummy
variables for $M$.
Under standard OLS assumptions, precision weights minimize the
standard error of the weighted mean estimator.

% Effect estimates may be improved by combining matching with regression
% adjustment
% \citep[e.g.]{rubin2000combining,rosenbaum2002covariance,ho:etal:2007}.
% Regression can improve matching estimators both by correcting for
% violations of Matched Ignorability due to observed covariate
% imbalances within matched sets (hence reducing the estimate's bias),
% and by explaining a portion of the variance in $Y$ and hence reducing
% the estimate's standard error.
% Here we use the regression adjustment method described in
% \citet{hansen2009b}, which takes place in three steps: (1) regressing posttest scores for students with
% $H=0$ on indicators for racial categories, grade, ESL status, economic disadvantage, and sex,
% and with a natural spline with five degrees of freedom for pretest,
% yielding a vector of coefficient estimates $\bm{\hat{\beta}}$;
% (2) computing residuals $e_i=Y_i-\bm{\hat{\beta}}^T\bm{\tilde{x}}_i$,
% where $\tilde{x}_i$ is the vector of covariates for subject $i$,
% for every unit $i$ in the sample; and (3) estimating effects in
% equation \eqref{eq:tauWeightedMean} with
% $\hat{tau}_m=\bar{e}_{M=m,H=1}-\bar{e}_{M=m,H=0}$, that is, with $e$
% in place of $Y$.

Table \ref{tab:matchResults} gives four treatment effect estimates
using these methods.
The first row gives the result using precision-weights without
regression adjustment. estimating an effect of $H$
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with a natural spline with five degrees of freedom for pretest.
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}.
All three sets of standard errors
and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator,
\citealt{sandwichPackage}.

The three estimates all roughly agree that being a high hint user
decreases posttest scores by roughly a fifth of a standard deviation.
That is, requesting hints hurts posttest scores---teachers should
recommend against students' requesting hints.

The similarity of the three estimates boosts our confidence, if only
slightly. The ``ATE'' and ``TOT'' estimates refer to different groups
of students---ATE estimates an average effect for the entire treatment
group, whereas TOT estimates an effect for high hint users. Had these
been very different, we might wonder why the effect of hint requests
varies so much between high and low hint users, and if that suggests
selection bias.
The fact that the additional covariate adjustment of the ANCOVA model
made little difference in the effect estimate suggests that the
propensity score match eliminated all or most of the confounding bias due to
measured covariates; hence, additional adjustment was unnecessary.

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Tue Dec 17 17:28:44 2019
\begin{table}[ht]
\centering
\begin{tabular}{rllccc}
  \hline
Weights & Estimate & Std. Error & CI & SI1 & SI2 \\ 
  \hline
OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.31,-0.04] \\ 
  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.31,0.06] \\ 
  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.33,0.05] \\ 
   \hline
\end{tabular}
\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
the causal question.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based (see Section
\ref{sec:dichotomizingHintUsage}),
we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group 70\% of students have $H=1$ and 30\% have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
individual level, as $\delta(1)_i$ and $\xi(0)_i$.
However, as in the usual causal inference case, individual effects are
not identified without strong assumptions, and the target of
estimation is instead taken to be super-population average effects,
$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$. Our approach to estimating these
is to estimate the means of each of the component potential outcomes,
$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
Y(0,H(0))=\EE Y(Z=0)$ are non-parametrically identified due to the
randomization of treatment assignment.

On the other hand, randomization alone does not identify the strictly counterfactual
potential outcome $\EE Y(1,H(0))$.
In a typical mediation analysis of an RCT, identification of the
distribution of $H(0)$ may be estimated based on the observed
distribution of $H$ in the control group, due to randomization.
In contrast, in the CTA1 study, $H$ is not observed in the control
group.
However, $H(0)$ was determined by the experimental design: since the
CTA1 software was unavailable to the control group, $H(0)=0$ for all
experimental participants, and $Y(1,H(0))\equiv Y(1,0)$.

For treated subjects $i$ for whom $H_i=0$, then, treatment assignment
did not affect $H$---$H_i(0)=H_i(1)=0$.
In other words, for these subjects,
\begin{equation*}
Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
\end{equation*}
Therefore, when $Z_i=1$ and $H_i=0$,
$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
indirect effect.

Treated subjects with $H=1$ present more of a challenge.
To estimate indirect effects for these subjects, we may rely on the
same causal setup as in the observational approach:
matched ignorability and the matching design of section
\ref{sec:observational}.
Each member of the treatment group with $H=1$ is matched to other
treatment-group subjects with $H=0$.
Matched ignorability states that within these matched sets, $Y(1,0)$
is independent of $H$.
Therefore, if subject $i$ is matched to subject $j$, $H_i=1$, and
$H_j=0$, we may estimate $Y_i(1,0)$ based on the observed
$Y_j=Y_j(1,0)$.
When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
$Y_i(1,0)$ is estimated with the average of their observed outcomes.

Specifically, let
\begin{equation}\label{eq:estY10}
\hat{Y}_i(1,0)=\begin{cases}
 Y_i & H_i=0\\[3ex]
\frac{\displaystyle\sum_{j:M_j=M_i}
  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
\end{cases}
\end{equation}
For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
average outcome of their matched pairs with $H=0$.

Now, for each member $i$ of the treatment group we have observed
$Y_i(1,H(1))=Y_i$, and estimated or inferred
$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
Then, we can estimate $\Delta$, the average indirect effect in the
treatment group, by averaging over the treatment group:
\begin{equation*}
 \hat{\Delta}\equiv\widehat{\EE[\delta(0)|Z=1]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
\end{equation*}
(Typically, the overall average indirect effect, $\EE[\delta(0)]$, is
estimated instead of the average indirect effect of the treatment
group, $\EE[\delta(0)|Z=1]$.
We estimated the latter since data from the control group was not used
in our estimation strategy, and estimating the overall average would
simply complicate our standard error estimation.)

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the average difference between the realized outcomes for students in
with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]
\end{equation*}

$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1,H(1))-Y(1,0)|Z=1]
\end{equation*}
This can be re-written:
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
\end{split}
\\[2ex]
&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)
&=Pr(H=1)TOT
\end{align*}
Since $Pr(H=1)=$0.7,
$\hat{\Delta}=$0.7$TOT$.
Estimated thusly, the indirect effect is negative: -0.1, with
a standard error of 0.04.
This analysis suggests that the availability of hints
actually \emph{lowers} the treatment effect, that CTA1 works not
because of hints, but despite them.
